Baseline Drift Estimation for Air Quality Data Using Quantile Trend Filtering

### 2019-04-11 Manuscript/detrendify_aoas.tex
More minor edits.

\begin{abstract}
	We address the problem of estimating smoothly varying baseline trends in time series data. This problem arises in several applications, including chemistry, macroeconomics, and medicine; however, our study is motivated by the analysis of data from low cost air quality sensors. Our methods extend the quantile trend filtering framework to allow the estimation of multiple quantile trends simultaneously while ensuring that the quantiles do not cross. To handle the computational challenge posed by very long time series, we propose a parallelizable alternating direction method of moments (ADMM) algorithm. The ADMM algorthim enables the estimation of trends in a piecewise manner, both reducing the computation time and extending the limits of the method to larger data sizes. We also address smoothing parameter selection and propose a modified criterion based on the extended Bayesian Information Criterion. Through simulation studies and our motivating application to low cost air quality sensor data, we demonstrate that our model provides better quantile trend estimates than existing methods and improves signal classification of low-cost air quality sensor output.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. Due to changes in temperature and relative humidity the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, which is illustrated by the three sharp transient spikes at 11:32, 14:10, and 16:03. However, the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed vector into a slowly varying component and a transient component can lead to greatly improved and simplified downstream analysis.

In a wide range of applications that spans chemistry \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{brantley2014mobile}, and medical sciences \citep{pettersson2013algorithm, marandi2015qualitative}, scalar functions of time $y(t)$ are observed and assumed to be a superposition of an underlying slowly varying baseline trend $\theta(t)$, other more rapidly varying components $s(t)$, and noise. In practice, $y(t)$ is observed at discrete time points $t_1, \ldots, t_n$, and we model the vector of signal samples $\VE{y}{i} = y(t_i)$ as
\begin{eqnarray*}
	\V{y} & = & \V{\theta} + \V{s} + \V{\varepsilon},
\end{eqnarray*}
where $\VE{\theta}{i} = \theta(t_i)$, $\VE{s}{i} = s(t_i)$, and $\V{\varepsilon} \in \Real^n$ is a vector of uncorrelated noise. For notational simplicity, for the rest of the paper, we assume that the time points take on the values $t_i = i$, but it is straightforward to generalize to an arbitrary grid of time points.

In some applications, the slowly varying component $\V{\theta}$ is the signal of interest, and the transient component $\V{s}$ is a vector of nuisance parameters. In our air quality application, the roles of $\V{\theta}$ and $\V{s}$ are reversed; $\V{s}$ represents the signal of interest and $\V{\theta}$ represents a baseline drift that obscures the identification of the important transient events encoded in $\V{s}$. We describe this motivating problem further to illustrate the critical importance of being able to separate $\V{s}$ from $\V{\theta}$.

@@

To tackle demixing problems, we introduce a scalable baseline estimation framework by building on \textit{$\ell_1$-trend filtering}, a relatively new nonparametric estimation framework. Our contributions are three-fold.
\begin{itemize}
	\item \cite{Kim2009} proposed using the check function as a possible extension of $\ell_1$-trend filtering but did not investigate it further. Here, we develop the basic $\ell_1$-quantile-trend-filtering framework and extend it to model multiple quantiles simultaneously with non-crossing constraints to ensure validity and improve trend estimates.
	\item To reduce computation time and extend the method to long time series, we develop a parallelizable ADMM algorithm. The algorithm proceeds by splitting the time domain into overlapping windows, fitting the model separately for each of the windows and reconciling estimates from the overlapping intervals.
	\item Finally, we propose a modified criterion for performing model selection.
\end{itemize}

@@

 As the size of the data increases, computation time becomes prohibitive \Fig{timing}. In our application to air quality sensor data, measurements are recorded every second resulting in 86,400 observations per day. This number of observations is already too large to use with currently available R packages used for estimating quantile trends \citep{fields, quantreg}. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose a divide-and-conquer approach via an ADMM algorithm for solving large problems in a piecewise fashion.

>>>>

\begin{abstract}
	We address the problem of estimating smoothly varying baseline trends in time series data. This problem arises in a wide range of fields, including chemistry, macroeconomics, and medicine; however, our study is motivated by the analysis of data from low cost air quality sensors. Our methods extend the quantile trend filtering framework to enable the estimation of multiple quantile trends simultaneously while ensuring that the quantiles do not cross. To handle the computational challenge posed by very long time series, we propose a parallelizable alternating direction method of moments (ADMM) algorithm. The ADMM algorthim enables the estimation of trends in a piecewise manner, both reducing the computation time and extending the limits of the method to larger data sizes. We also address smoothing parameter selection and propose a modified criterion based on the extended Bayesian Information Criterion. Through simulation studies and our motivating application to low cost air quality sensor data, we demonstrate that our model provides better quantile trend estimates than existing methods and improves signal classification of low-cost air quality sensor output.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. Due to changes in temperature and relative humidity the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, which is illustrated by the three sharp transient spikes at 11:32, 14:10, and 16:03. However, the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed time series into a slowly varying component and a transient component can lead to greatly improved and simplified downstream analysis.

While this work is motivated by the analysis of data from low cost air quality sensors, the problem of demixing nosy time series into trends and transients is ubiquitous across many fields of study. In a wide range of applications that spans chemistry \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{brantley2014mobile}, and medical sciences \citep{pettersson2013algorithm, marandi2015qualitative}, scalar functions of time $y(t)$ are observed and assumed to be a superposition of an underlying slowly varying baseline trend $\theta(t)$, other more rapidly varying components $s(t)$, and noise. In practice, $y(t)$ is observed at discrete time points $t_1, \ldots, t_n$, and we model the vector of signal samples $\VE{y}{i} = y(t_i)$ as
\begin{eqnarray*}
	\V{y} & = & \V{\theta} + \V{s} + \V{\varepsilon},
\end{eqnarray*}
where $\VE{\theta}{i} = \theta(t_i)$, $\VE{s}{i} = s(t_i)$, and $\V{\varepsilon} \in \Real^n$ is a vector of uncorrelated noise. For notational simplicity, for the rest of the paper, we assume that the time points take on the values $t_i = i$, but it is straightforward to generalize to an arbitrary grid of time points.

In some applications, the slowly varying component $\V{\theta}$ is the signal of interest, and the transient component $\V{s}$ is a vector of nuisance parameters. In our air quality application, the roles of $\V{\theta}$ and $\V{s}$ are reversed; $\V{s}$ represents the signal of interest and $\V{\theta}$ represents a baseline drift that obscures the identification of the important transient events encoded in $\V{s}$.

@@

To tackle demixing problems, we introduce a scalable baseline estimation framework by building on \textit{$\ell_1$-trend filtering}, a relatively new nonparametric estimation framework. Our contributions are three-fold.
\begin{itemize}
	\item \cite{Kim2009} proposed using the check function as a possible extension of $\ell_1$-trend filtering but did not investigate it further. Here, we develop the basic $\ell_1$-quantile-trend-filtering framework and extend it to model {\em multiple quantiles simultaneously with non-crossing constraints to ensure validity and improve trend estimates}.
	\item To reduce computation time and extend the method to long time series, we develop a parallelizable ADMM algorithm. The algorithm proceeds by splitting the time domain into overlapping windows, fitting the model separately for each of the windows and reconciling estimates from the overlapping intervals.
	\item Finally, we propose a modified criterion for performing model selection.
\end{itemize}

@@

 As the size of the data increases, computation time becomes prohibitive (\Fig{timing}). In our application to air quality sensor data, measurements are recorded every second resulting in 86,400 observations per day. This number of observations is already too large to use with currently available R packages used for estimating quantile trends \citep{fields, quantreg}. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose a divide-and-conquer approach via an ADMM algorithm for solving large problems in a piecewise fashion.
